---
title: An R Markdown document converted from "1_NTU_ab_to_pseq.ipynb"
output: html_document
---


### Analysis of microbial communities from totalRNA data processed using phyloFlash. ###

```{r}
# Load necessary libraries
library(dplyr)
library(microbiome)
library(microViz)
library(viridis)
library(vegan)
library(here)
library(tibble)

library(readr)
library(tidyr)
library(purrr)
library(stringr)
```

### Scale Function to Normalize Library Sizes

This function, `scale_reads`, normalizes the read counts in a phyloseq object to a specified library size `n`. It performs the following steps:

1. **Transforms Counts**: Scales the read counts in each sample so that the total reads per sample equals `n`.
2. **Rounds Counts**: Floors the scaled counts to convert them back to integers.
3. **Removes Zero-Abundance Taxa**: Drops any taxa that have zero total abundance across all samples after scaling.

This is useful for reducing library size bias while retaining proportional abundance information.

```{r}
### scale function to normalize library sizes
scale_reads <- function(physeq, n) {
  physeq.scale <-
    transform_sample_counts(physeq, function(x) {
      (n * x/sum(x))
    })
  otu_table(physeq.scale) <- floor(otu_table(physeq.scale))
  physeq.scale <- prune_taxa(taxa_sums(physeq.scale) > 0, physeq.scale)
  return(physeq.scale)
}
```

```{r}
# --- Taxonomic fallback filler ---
fill_unclassified_clean <- function(df) {
  for (i in 2:ncol(df)) {
    for (j in 1:nrow(df)) {
      if (is.na(df[j, i])) {
        last_non_na <- NA
        for (k in (i - 1):1) {
          val <- df[j, k]
          if (!is.na(val) && !grepl("^Unclassified", val)) {
            last_non_na <- val
            break
          }
        }
        if (!is.na(last_non_na)) {
          df[j, i] <- paste("Unclassified", last_non_na)
        }
      }
    }
  }
  return(df)
}

# --- SILVAtaxopath function ---
SILVAtaxopath <- function(tax, SILVA) {
  rank_order <- c(
    "domain", "superkingdom", "kingdom", "subkingdom", "major_clade",
    "superphylum", "phylum", "subphylum", "infraphylum",
    "superclass", "class", "subclass", "infraclass",
    "superorder", "order", "suborder",
    "superfamily", "family", "subfamily", "genus"
  )
  available_ranks <- intersect(rank_order, levels(SILVA$rank))
  output <- matrix(NA, nrow = length(tax), ncol = length(available_ranks))
  colnames(output) <- available_ranks
  for (i in seq_along(tax)) {
    for (j in seq_along(tax[[i]])) {
      path_j <- paste(tax[[i]][1:j], collapse = ";")
      match_idx <- which(SILVA$path == path_j)
      if (length(match_idx) > 0) {
        rank_j <- as.character(SILVA$rank[match_idx])
        node_j <- as.character(SILVA$node[match_idx])
        if (rank_j %in% available_ranks) {
          output[i, rank_j] <- node_j
        }
      }
    }
  }
  return(output)
}
```

```{r}
# Step 1: List all .csv files
all_paths <- list.files(
  path = here("C:/Users/PC/Desktop/euk_phyloflash/phyloflash_NTUabundance_files"),
  pattern = "\\.csv$",
  full.names = TRUE
)

if (length(all_paths) == 0) stop("No CSV files found.")

# Step 2: Extract sample names
sample_names <- all_paths %>%
  basename() %>%
  sub("\\.phyloFlash\\.NTUabundance\\.csv$", "", .)

# Step 3: Read and label each file
all_content <- Map(function(file, sample) {
  df <- read_csv(file, col_names = c("Tax", sample), col_types = cols())
  return(df)
}, all_paths, sample_names)

# Step 4: Merge on taxonomy column
NTUtable <- reduce(all_content, full_join, by = "Tax")

# Preserve the "taxon" column name
new_colnames <- colnames(NTUtable)
sample_cols <- new_colnames[-1]  # exclude 'taxon'

# Rename each sample column from S2301C to S23_01_C
renamed_samples <- gsub("^S(\\d{2})(\\d{2})([A-Za-z])$", "S\\1_\\2_\\3", sample_cols)

# Combine with the unchanged first column name
colnames(NTUtable) <- c("Tax", renamed_samples)

# Step 5: Replace NAs with 0
NTUtable[is.na(NTUtable)] <- 0
```

```{r}
head(NTUtable)
```

```{r}
# --- Step 2: Run SILVAtaxopath for taxonomy parsing ---
taxpaths <- NTUtable$Tax
tax_split <- strsplit(taxpaths, ";")

silva_raw <- read.table("C:/Users/PC/Desktop/euk_phyloflash/tax_slv_ssu_138.2.txt", sep = "\t", header = FALSE, stringsAsFactors = FALSE)
silva_map <- data.frame(
  path = gsub(";$", "", silva_raw$V1),
  node = sapply(strsplit(silva_raw$V1, ";"), function(x) x[length(x)]),
  rank = silva_raw$V3,
  stringsAsFactors = TRUE
)

taxonomy_matrix <- SILVAtaxopath(tax_split, silva_map)
taxonomy_df <- as.data.frame(taxonomy_matrix, stringsAsFactors = FALSE)

# --- Step 3: Remove unwanted groups ---
taxonomy_df_filtered <- taxonomy_df %>%
  filter(
    domain %in% c("Bacteria", "Archaea"),
    if_any(everything(), ~ !(. %in% c("Chloroplast", "Mitochondria")))
  )

# --- Step 4: Apply unclassified fallback logic ---
taxonomy_filled <- fill_unclassified_clean(taxonomy_df_filtered)

# --- Step 5: Match taxonomy_filled back to NTUtable (by row) ---
NTUtable_filtered <- NTUtable[rownames(taxonomy_df_filtered), ]

# --- Step 6: Combine final taxonomy + abundance table ---
final_16S_table <- bind_cols(taxonomy_filled, NTUtable_filtered[, -1])  # remove raw "Tax" column

# Step 7: Add OTU ID
otu_ids <- paste0("OTU", seq_len(nrow(taxonomy_filled)))

# Step 8: Select only desired taxonomy columns
desired_tax_cols <- c("domain", "major_clade", "kingdom", "phylum", "class", "order", "family", "genus")
taxonomy_selected <- taxonomy_filled[, desired_tax_cols, drop = FALSE]

# Step 8.5: Rename taxonomy columns to desired format
colnames(taxonomy_selected) <- c(
  "Domain", "Major Clade", "Kingdom", "Phylum", "Class", 
  "Order", "Family", "Genus"
)

# Step 8.6: Add Species column
taxonomy_selected$Species <- ifelse(
  !is.na(taxonomy_selected$Genus) & !grepl("^Unclassified", taxonomy_selected$Genus),
  paste("Unclassified", taxonomy_selected$Genus),
  "Unclassified"
)

# Step 9: Combine final taxonomy + abundance + OTU ID
final_16S_table <- bind_cols(data.frame(otu_id = otu_ids), taxonomy_selected, NTUtable_filtered[, -1])

# --- Step 8: Done ---
head(final_16S_table)
```

```{r}
write.csv(final_16S_table, file="NTU_combined_prok.csv")
```

```{r}
taxonomy_short = c("Domain", "Major Clade", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species")

tax <- final_16S_table %>%
  select(all_of(taxonomy_short), otu_id) %>%
  column_to_rownames(var = "otu_id") %>%
  as.matrix()

otu <- final_16S_table %>%
  select(-all_of(taxonomy_short)) %>%
  column_to_rownames(var = "otu_id") %>%
  as.matrix()
```

```{r}
head(tax)
```

```{r}
head(otu)
```

```{r}
#create sample data
SAMPLE_table <- read.csv(file = here("C:/Users/PC/Desktop/euk_phyloflash/metadata.csv"), header = TRUE, row.names = 1,sep = "\t")
SAMPLE_DATA <- sample_data(SAMPLE_table)

# Extract, transform, and reassign row names
rownames(SAMPLE_DATA) <- gsub(
  "^S(\\d{2})(\\d{2})([A-Za-z])$", "S\\1_\\2_\\3",
  rownames(SAMPLE_DATA)
)
```

```{r}
SAMPLE_DATA
```

```{r}
# Check OTU table names vs mapping file names (if they don't match, samples won't be included in final phyloseq object)
all(colnames(otu) %in% rownames(SAMPLE_DATA))
```

```{r}
PHYSEQ <- phyloseq(
  otu_table(otu, taxa_are_rows = TRUE),
  tax_table(tax),
  SAMPLE_DATA
)
```

```{r}
sample_data(PHYSEQ)
```

```{r}
pseq_clean <- PHYSEQ 
```

```{r}
save(pseq_clean, file = "./Fixed_ps.RData")

otu_table(pseq_clean) %>%
  as.data.frame() %>%
  rownames_to_column("id") %>%
  left_join(tax_table(pseq_clean) %>% as.data.frame() %>% rownames_to_column("id")) %>%
  write.csv(file = here("phyloseq_biom_16S.csv"), row.names = FALSE)
```
library(microeco)

saveRDS(pseq_clean, file = "16S_phyloflash.rds")




